Follow-up Samples

### 2017-09-13 submission/manuscript_R2.Rmd
bring abstrct into word count for msphere

**Background.** Colorectal cancer is a worldwide health problem. Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma. This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon. We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).

**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups. Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment. To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples. Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival. Although it was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma, this was not the case for carcinomas. Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.

### Keywords

microbiota; colorectal cancer; polyps; treatment; risk factor.

>>>>

**Background.** Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma. This study tests whether treatment alters the abundance of bacterial populations associated with disease versus those with a normal colon by sequencing the 16S rRNA genes in feces of `r length(rownames(good_metaf))` individuals before and after treatment (adenoma n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma n = `r change_summary["thetayc", "srn_n"]`), and carcinoma n = `r change_summary["thetayc", "crc_n"]`).

**Results.** The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although Random Forest models classified pre and post-treatment samples for each diagnosis group, there
was little overlap between the OTUs that were indicative of each treatment. Training Random Forest models on a larger cohort that contained individuals with normal colons, adenomas, advanced adenomas, and carcinomas we assessed the positive probability of having one of three diagnoses and whether the pre and post-treatment samples were classified the same. Patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). There was no significant difference in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** It was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma. However, for carcinomas, not only were there large changes in pre versus post-treatment samples, but these changes were towards a more normal microbiota.

### 2017-09-13 submission/manuscript_R2.tex
bring abstrct into word count for msphere

\textbf{Background.} Colorectal cancer is a worldwide health problem.
Despite growing evidence that members of the gut microbiota can drive
tumorigenesis, little is known about what happens to it after treatment
for an adenoma or carcinoma. This study tested the hypothesis that
treatment for adenoma or carcinoma alters the abundance of bacterial
populations associated with disease to those associated with a normal
colon. We tested this hypothesis by sequencing the 16S rRNA genes in the
feces of 67 individuals before and after treatment for adenoma (N = 22),
advanced adenoma (N = 19), and carcinoma (N = 26).

\textbf{Results.} There were small changes to the bacterial community
associated with adenoma or advanced adenoma and large changes associated
with carcinoma. The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although treatment was associated with intrapersonal
changes, the change in the abundance of individual OTUs in response to
treatment was not consistent within diagnosis groups (P-value
\textgreater{} 0.05). Because the distribution of OTUs across patients
and diagnosis groups was irregular, we used the Random Forest machine
learning algorithm to identify groups of OTUs that could be used to
classify pre and post-treatment samples for each of the diagnosis
groups. Although the three models could differentiate between the pre
and post-treatment samples, there was little overlap between the OTUs
that were indicative of each treatment. To determine whether individuals
who underwent treatment were more likely to have OTUs associated with
normal colons we used a larger cohort that contained individuals with
normal colons and those with adenomas, advanced adenomas, and
carcinomas. We again built Random Forest models and measured the change
in the positive probability of having one of the three diagnoses to
assess whether the post-treatment samples received the same
classification as the pre-treatment samples. Samples from patients who
had carcinomas changed towards a microbial milieu that resembles the
normal colon after treatment (P-value \textless{} 0.05). Finally, we
were unable to detect any significant differences in the microbiota of
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} By better understanding the response of the
microbiota to treatment for adenomas and carcinomas, it is likely that
biomarkers will eventually be validated that can be used to quantify the
risk of recurrence and the likelihood of survival. Although it was
difficult to identify significant differences between pre and
post-treatment samples from patients with adenoma and advanced adenoma,
this was not the case for carcinomas. Not only were there large changes
in pre versus post-treatment samples for those with carcinoma, but these
changes were towards a more normal microbiota.

\subsubsection{Keywords}\label{keywords}

microbiota; colorectal cancer; polyps; treatment; risk factor.

>>>>

\textbf{Background.} Despite growing evidence that members of the gut
microbiota can drive tumorigenesis, little is known about what happens
to it after treatment for an adenoma or carcinoma. This study tests
whether treatment alters the abundance of bacterial populations
associated with disease versus those with a normal colon by sequencing
the 16S rRNA genes in feces of 67 individuals before and after treatment
(adenoma n = 22), advanced adenoma n = 19), and carcinoma n = 26).

\textbf{Results.} The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although Random Forest models classified pre and
post-treatment samples for each diagnosis group, there was little
overlap between the OTUs that were indicative of each treatment.
Training Random Forest models on a larger cohort that contained
individuals with normal colons, adenomas, advanced adenomas, and
carcinomas we assessed the positive probability of having one of three
diagnoses and whether the pre and post-treatment samples were classified
the same. Patients who had carcinomas changed towards a microbial milieu
that resembles the normal colon after treatment (P-value \textless{}
0.05). There was no significant difference in the microbiota of
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} It was difficult to identify significant
differences between pre and post-treatment samples from patients with
adenoma and advanced adenoma. However, for carcinomas, not only were
there large changes in pre versus post-treatment samples, but these
changes were towards a more normal microbiota.
